---
title: "Mini Data-Analysis Deliverable 1"
output: html_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators. A link to help you create a private repository is available on the #collaborative-project Slack channel. 

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: building_permits
2: steam_games
3: vancouver_trees
4: flow_sample

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

```{r}
### EXPLORE HERE ###
glimpse(building_permits)
class(building_permits)
dim(building_permits)
#Confirmed tibble with 20680 rows and 14 columns. 

head(steam_games)
class(steam_games)
dim(steam_games)
#confirmed tibble with 40833 rows and 21 columns. Noticed multiple answers inn popular_tags variable and long descriptions

glimpse(vancouver_trees)
class(vancouver_trees)
dim(vancouver_trees)
#Confirmed tibble with 14661 rows and 20 columns. Straightforward data and variables can be investigated in relation to another.

head(flow_sample)
glimpse(flow_sample)
dim(flow_sample)
class(flow_sample)
#Confirmed tibble with 218 rows and 7 columns. Smallest dataset, nots of numeric variables.
```

<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below ---------------------------->
2 datasets I am cosidering now is steam_games and vancouver_trees because the data is straightforward and I am able to think of relationships between the different variables. Additionally, these two datasets have fewer variables to work with so it will be easy to work with.
<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below ---------------------------->
For steam_games, I would be interested to see the relationship between release_date and genre to see if there was a specific time period where certain genres were more popular. I could plot the relationship by bar graph.

For vancouver_trees, I would be interested manipulating diameter to find the average diameter and neighbourhood_name to see which neighbourhoods have on average wider diameter trees. 

My final choice is vancouver_trees because I'm interested to know more about the landscape in Vancouver which is the city that I was born and raised in.
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->

<!----------------------------------------------------------------------------->

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->
**Task 2: I will be doing number 1,4,5, and 6**
1. Plot the distribution of a numeric variable. I will be looking at the distribution of the height_range_id to see which height ranges are most common in Vancouver.
```{r}
ggplot(vancouver_trees, aes(height_range_id)) + geom_histogram(bins = 15)
#I played around with different bins commands and found that 15 gave the nicest graph. 
```

4. Explore the relationship between 2 variables in a plot. I will be looking to see if there are any trends in the height range of trees in Vancouver as time goes on.
```{r}
ggplot(vancouver_trees, aes(date_planted, height_range_id)) + geom_line(colour = "blue")
#After looking at this plot, I noticed that the the trees planted between 2010-2020 have shorter height ranges than trees in 2000-2010.
```

5. Filter observations in your data according to your own criteria. I will be looking at diameter and see how I want to filter the variable.
```{r}
max(vancouver_trees$diameter) #435
min(vancouver_trees$diameter) #0
median(vancouver_trees$diameter) #9
#After knowing the max, min, and median values of diameter, I was surprised to know that the median was 9 which is so much closer to the min than max so I want to filter all the trees that have a diameter less than 9 to see what is going on.

filter(vancouver_trees, diameter < 9) 
```
6. Use a boxplot to look at the frequency of different observations within a single variable. I will be looking at the diameter of the tree by whether they do or do not have a root barrier.
```{r}
ggplot(vancouver_trees, aes(root_barrier, diameter)) + geom_boxplot()
#After looking at this boxplot, it looks like trees with a root barrier have smaller diameters than trees without root barriers.
```

**Task 3: Write your research questions**
1. What are the average diameter of trees in each neighbourhood in Vancouver?
2. Does having a root barrier result in a smaller tree diameter? 
3. Does having a root barrier result in a smaller height range tree?
4. Is the height range of trees getting smaller as the years go on?


# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.
8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
**Research Question 1: What are the average diameter of trees in each neighbourhood in Vancouver?**

*Summarize 1*
```{r}
#To find the average mean of trees in each neighbourhood
neighbourhood_meandiameter <- vancouver_trees %>%
  group_by(neighbourhood_name) %>%
  summarise(diameter_mean = mean(diameter, na.rm = TRUE)) 

#Now I want to know the summary statistics of the average diameter between the different neighbourhoods.
#Making sure I have the dataset I want
glimpse(neighbourhood_meandiameter)
#Finding range
range(neighbourhood_meandiameter$diameter_mean)
#Can determine min and max for range aas well as mean, median, 1st and 3rd quartile
summary(neighbourhood_meandiameter) 
#Finding standard deviation
sd(neighbourhood_meandiameter$diameter_mean)
```
From the summary or range function, the range of average diameters grouped by different neighbourhoods is 7.446 - 14.240. From the summary function, the mean is 11.435 and the median is 11.435. The standard deviation is 1.684479.

*Graphing 8*
```{r}
ggplot(neighbourhood_meandiameter, aes(diameter_mean)) + geom_histogram(bins = 10)
ggplot(neighbourhood_meandiameter, aes(diameter_mean)) + geom_histogram(bins = 20)
ggplot(neighbourhood_meandiameter, aes(diameter_mean)) + geom_histogram(bins = 40)
```
From the different histograms, I plotted where bins = 10, 20, and 40 and I think the bins = 10 looks the best because you can see the distribution better and there are no big gaps between the different bins.

**Research question 2: Does having a root barrier result in a smaller tree diameter?**

*Summarize 1*
```{r}
#First I want to compare statistics between trees that having a root barrier and trees that do not
rootbarrier_y <- filter(vancouver_trees, root_barrier == "Y")
glimpse(rootbarrier_y)

rootbarrier_n <- filter(vancouver_trees, root_barrier =="N")
glimpse(rootbarrier_n)

#Now I will look at range, mean, standard deviation, and median of diameter for both datasubsets. 
summary(rootbarrier_y$diameter)
sd(rootbarrier_y$diameter)
#Min = 0.5, Max = 86, Median = 3.0, Mean = 4.4, SD = 2.998134

summary(rootbarrier_n$diameter)
sd(rootbarrier_n$diameter)
#Min = 0, Max = 435, Median = 10.0, Mean = 11.96, SD = 9.290934
```
Comparing the statistics between the diameter of trees that do or do not have a root barrier shows that trees with a root barrier on average is smaller than trees that do not have a root barrier.

*Graphing 6*
```{r}
ggplot(vancouver_trees, aes(root_barrier, diameter)) + geom_jitter(alpha = 0.5) + xlab("Root Barrier") 
```
This jitterplot visualizes that trees trees with a root barrier have a smaller diameter compared to trees without a root barrier. This is a good time to make the transparency lower so that the dots on the jitterplot are not too crowded.

**Research Question 3: Does having a root barrier result in a shorter height tree?**

*Summarize 2*
```{r}
#To count the number of observations, I will be grouping the height ranges into levels and counting how many trees are in each level.
rootbarrier_y$height_range_id <- as.factor(rootbarrier_y$height_range_id)
summary(rootbarrier_y$height_range_id)
#Level 0 = 43,    1 = 6998,   2 = 1633,    3 = 314,    4 = 142,    5 = 13    6 = 8,    7 = 2,    8 = 2,    9 = 1

rootbarrier_n$height_range_id <-as.factor(rootbarrier_n$height_range_id)
summary(rootbarrier_n$height_range_id)
#Level 0 = 171, 1 = 32961, 2 = 40573, 3 = 25993, 4 = 20388, 5 = 9001, 6 = 5187, 7 = 2223, 8 = 744, 9 = 202, 10 = 12
```
Comparing the number of observations in higher level height ranges between trees that do and do not have root barriers, you can see that trees that do not have root barriers have more observations seen in the heigher level of height ranges.

*Graphing 
```{r}
ggplot(vancouver_trees, aes(root_barrier, height_range_id)) + geom_jitter(colour = "dark green", alpha = 0.5) + xlab("Root Barrier") + ylab("Height Range")
```
This jitterplot shows that there are more trees in different height ranges if they do not have a root barrier.

**Research Question 4: Is the height ranges of trees getting smaller as the years go on?**

*Summarize 1*
```{r}
summary(vancouver_trees$height_range_id)
sd(vancouver_trees$height_range_id)
```
From these two functions, I can see the range of height-ranges are from 0 - 10. The mean is 2.627, the median is 2 and the standard deviation is 1.542994.

*Graphing 5*
```{r}
ggplot(vancouver_trees, aes(date_planted, height_range_id)) + geom_line(colour = "Blue") + geom_point(colour = "blue", size = 0.5)
```
From this graph, you can see that there were trees with higher height range before 2000 and the height range slowly declines from 2000 to 2020. This graph has geom_point added onto geom_line (2 layers) so that you can still the full range of heights. 
<!----------------------------------------------------------------------------->

### 1.2 (3 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below ---------------------------->
Based on these exercises, I think I can close to answering question 2-4. Questions 2 and 2 where I look at the impact of root barrier on height and diameter shows that having a root barrier negatively impacts these two qualities of a tree. In question 4, the graph visualizes very well that the height of trees after 2000 have a negative trend. For question 1, the summarize exercise is able to tell me the average diameter of trees in each neighbourhood but the plot exercise does not contribute much to answering the research question.

<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.